Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence

### 2019-03-22 eDriversMSv2.rmd
genesis

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Hugo Bourdages^7^, Clément Chion^6^, Peter S. Galbraith^7^, Benjamin Halpern^8,9,10^, Camille Lavoie^2^, Christopher W. McKindsey^7^, Alfonso Mucci^11^, Simon Pineault^12^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
date: 2019-02-06
fontsize: 12pt
output:
  pdf_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/eDrivers/eDriversMS/
rmarkdown::render('./eDriversMS.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS.rmd', 'word_document')
-->

@@

<!-- Setting up parameters for code snippet presented in Box 1 -->
\lstset{language=R,
    basicstyle=\small\ttfamily,
    stringstyle=\color{DarkGreen},
    otherkeywords={fetchDrivers, install_github, getData, kde2d},
    morekeywords={TRUE,FALSE},
    deletekeywords={data,frame,length,as,character},
  keywordstyle=\color{blue},
  commentstyle=\color{DarkGreen},
}

^1^Institut des sciences de la mer, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^2^Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
^4^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^5^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
^6^Institut des sciences de la forêt tempérée, Université du Québec en Outaouais, Gatineau, QC, Canada \newline
^7^Fisheries and Oceans Canada, Maurice Lamontagne Institute, Mont-Joli, QC, Canada \newline
^8^National Center for Ecological Analysis and Synthesis, University of California, Santa Barbara, CA, United States \newline
^9^Bren School of Environmental Science and Management, University of California, Santa Barbara, CA, United States \newline
^10^Imperial College London, Ascot, United Kingdom \newline
^11^Department of Earth & Planetary Sciences, McGill University, Montréal, QC, Canada \newline
^12^Ministère Environnement et Lutte contre les changements climatiques, Québec, QC, Canada  \newline

**Correspondence**: \newline
David Beauchesne \newline
david.beauchesne@uqar.com \newline

**Number of figures**: 5 \newline
**Number of tables**: 1 \newline

@@

# Abstract

Gathering environmental data for large scale, systematic initiatives such as
ecosystem-based management is a very challenging process. Building such
holistic datasets is made increasingly possible through initiatives like the
Ocean Biogeographic Information System (OBIS) and Bio-ORACLE. Data on
environmental drivers (*e.g.* climate change and fisheries) are however largely
missing even though their characterization is crucial to a better understanding
of the effects of multiple environmental threats. In this paper we discuss how
we are addressing this issue by launching an open data platform called *eDrivers*, focused
on characterizing and sharing data on the distribution and intensity of drivers
and using the Estuary and Gulf of St. Lawrence in eastern Canada as a case study.
The core objective underlying this initiative is to create a community-led
project built on a series of guiding principles aiming at upholding the highest
standards of data management and open science. The platform currently shares data for 22 drivers
including coastal, climate, fisheries and marine traffic related layers. Using
the currently available data, we show that few areas are free of cumulative
exposure in the St. Lawrence. The Estuary, the Anticosti Gyre and coastal areas
are particularly exposed, especially in the vicinity of urban centers. We also
identify 7 distinct areas of similar cumulative exposure regime (*i.e.* threat
complexes) that show that numerous drivers typically co-occur in different
regions of the St. Lawrence and that coastal areas are exposed to all driver
types. These observations are destined to improve as *eDrivers* evolves
through time to address gaps in knowledge, incorporates additional driver
layers and refines the characterization of current driver layers. Ultimately,
we believe that *eDrivers* represents a much needed solution that could
radically influence broad scale research and management practices by increasing
data accessibility and interoperability and by increasing research and
decision-making efficiency. All that is required is a commitment to open data,
adaptive monitoring and an integrated vision of ecosystem management.

**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
cumulative exposure, threat complex, multiple stressors, global change

# Introduction

@@

Gathering environmental data for large scale, systematic initiatives can however
be a very challenging and time consuming - not to say painful - process. On one
hand, there is an overwhelming and expanding wealth of data available. Such
information overload may reduce ability to take decision based on scientific
information, cause massive effort duplication, disproportionately appropriate
research funds for certain sectors, and obscure knowledge gaps amid a sea of information
[@eppler2004]. On the other hand, crucial data are lacking and remain largely
unavailable or inaccessible for a variety of reasons, including proprietary
rights, lack of organizational time, capacity and training, and unwillingness
to share, curtailing our ability for appropriate decision-making. When coupled
with political intricacies, such as multi-jurisdictional management, the data
gathering process can quickly become intractable.

Moreover, the current digital infrastructure is highly
decentralized and the practices are heterogeneous, preventing us from maximizing
benefits from research investments [@wilkinson2016]. Per @wilkinson2016, *"the
reason that we often need several weeks (or months) of specialist technical
effort to gather the data necessary to answer such research questions is not
the lack of appropriate technology; the reason is, that we do not pay our
valuable digital objects the careful attention they deserve when we create and
preserve them"*.

There now exists multiple initiatives that address this issue by assembling,
organizing and sharing vast arrays of environmental data.
Biotic data can be accessed through web portals such as the Ocean Biogeographic
Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
environmental data for specific areas also exist, such as the St. Lawrence
Global Observatory (SLGO; https://ogsl.ca/en), the European Marine Observation
and Data Network (EMODnet; http://www.emodnet.eu/) and the U.S. Integrated Ocean
Observing System (IOOS; https://ioos.noaa.gov/). Essential environmental
parameters are also organized, coordinated and acquired through global
initiatives like the Group on Earth Observations Biodiversity Observation
Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
and the Global Ocean Observing System (GOOS; http://www.goosocean.org).
However, environmental drivers, *i.e.* any externality that affects environmental processes
and disturbs natural systems and often referred to as stressors or pressures, remain
conspicuously missing from most of these initiatives [but see @halpern2015].
This is in spite of the need for high resolution data on their distribution and
intensity at a scale suitable for environmental management and impact assessments.

Drivers may be natural (e.g. sea surface
temperature anomalies and hypoxia) or anthropogenic (e.g. fisheries and marine
pollution) in origin.
The potential for complex interactions between drivers remains the largest
uncertainty when studying or predicting environmental change
[@darling2008; @cote2016]. The effects of multiple drivers can combine
non-linearly and result in effects that are greater (synergistic effect) or
lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016]. The uncertainty associated with complex
driver interactions must therefore be taken into account [@cote2016], yet most
research on driver effects in marine environments remains overwhelmingly focused
on single driver assessments [@obrien2019]. This is partly explained by the
complexity of studying multiple drivers simultaneously [@cote2016], but also
by the need for standardized data management practices and tools tailored to
the study of the effects of multiple drivers [@dafforn2016; @stock2018].

To address these issues, we are launching an initiative called *eDrivers*, an
open data platform focused on the description of the spatial distribution and
intensity of drivers and using the Estuary and Gulf of St. Lawrence in Canada
as a case study. The overarching goal of this platform is to create an expert
community committed to structuring, sharing and standardizing drivers data in
support to science and management.

The objective of this paper is twofold. The first is to present the guiding
principles underlying the inception and structure of *eDrivers*. We also wish
this paper to serve as a call for collaboration to any person, group or
organization who might be interested in contributing to this initiative. The
second objective focuses using the platform content to describe the intensity and distribution of drivers
in the St. Lawrence system and aims at 1) identifying areas of high cumulative exposure,
2) at characterizing areas with similar cumulative exposure regimes, dubbed threat
complexes [@bowler2019] and 3) at identifying drivers that are likely to interact
in the St. Lawrence.

@@

The St. Lawrence system, formed by one of the largest estuaries in the world and a
vast interior sea, is an complex social-ecological system characterized by highly
variable environmental conditions and oceanographic processes, both in space
and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique and
heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. It concomitantly provides
a wealth of ecosystem services that have historically and contemporarily
benefited the Canadian economy. It holds a rich fisheries industry targeting
over 50 species, serves as the gateway to eastern North-America by granting
access to more than 40 ports, has a booming tourism industry and an expanding
aquaculture production, develops emerging activities and has a yet untapped
hydrocarbon potential [@beauchesne2016; @schloss2017]. With major
investments recently made and more forthcoming in economic and infrastructure
development and research [*e.g.* @goq2015; @rqm2018], an intensification of the
human footprint is expected in the St. Lawrence system.

Arising from this expected intensification and the experience of past and
current ecological tragedies [*e.g.* the collapse of cod fisheries and the
decline of the beluga population; @frank2005; @plourde2014; @dempsey2018] is a
call for an integrative and holistic approach to manage the St. Lawrence as a
whole. Yet, jurisdiction over the St. Lawrence is divided between the Canadian
federal government and five provincial governments (Newfoundland and
Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec) with
each governmental entity further divided into multiple departments.
Researchers from dozens of academic institutions and
countless local organizations also share an interest in the St. Lawrence. This
wide range of actors results in highly scattered data management
and sharing processes, which we wish to address through this open data platform
initiative.

@@

Five guiding principles motivate the development of the platform and shape its
structure (Figure \ref{diag}).

@@

***Why***: Operating over such large scales in time, space and subject matter
requires a vast and diverse expertise that cannot possibly be possessed by any
one individual or organization. Consequently, we envision an initiative that
seeks to mobilize all individuals and entities with relevant expertise.

***How***: By promoting, consolidating and working with experts involved in
existing and highly valuable environmental initiatives already in place in the
St. Lawrence. Notable examples of environmental initiatives are the annual
review of physical [@galbraith2018], chemical, and biological [@blais2018]
oceanographic conditions in the St. Lawrence, the fisheries monitoring program
[@dfo2016], the annual groundfish and shrimp multidisciplinary survey
[@bourdages2018], the characterisation of benthic [@dutil2011], epipelagic and
coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
classification [@eccc2018]. There are also nascent efforts to make available
information on several human activities in the St. Lawrence such as the Marine
Spatial Data Infrastructure portal, which provides data on zoning, shipping,
port activities, and other human activities in Canadian waters, including the
St. Lawrence system [@goc2016].

By working with existing data portals whose objective is to share
environmental data. We are thus collaborating actively with the St. Lawrence
Global Observatory (SLGO) to develop the initiative and to host the
platform on their web portal (https://ogsl.ca/en). The mission of SLGO is to
promote and facilitate the accessibility, dissemination and exchange of official
and quality data and information on the St. Lawrence ecosystem through the
networking of organisations and data holders to meet their needs and those of
users, to improve knowledge and to assist decision-making in areas such as
public safety, climate change, transportation, resources and biodiversity
conservation. The SLGO is also one of three regional association spearheading
the Canadian Integrated Ocean Observing System
(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
which will focus on integrating oceanographic data from multiple sources to
make them accessible to end-users and to enable the national coordination of
ocean observing efforts by integrating isolated or inaccessible data, and by
identifying gaps or duplications in observations and research efforts.
We are also developing collaborations with the Portal on water knowledge
(http://www.environnement.gouv.qc.ca/eau/portail/), an initiative from the
Québec provincial government. This portal aims at collecting and sharing
accurate, complete and updated ressources on water and aquatic ecosystems to
support the mandate of relevant actors and stakeholders working in
water and aquatic ecosystems management.

By actively inviting, seeking, and developing collaborations as well as
encouraging constructive criticism from the inception and throughout the
lifetime of the platform.

By inviting external community contributions (Figure \ref{diag}).
External researchers or entities wishing to submit marine data will be able to
do so through the SLGO web portal (https://ogsl.ca/en). Submissions through
other data portals will also be accepted either through the development of data
sharing agreements or with the caveat that shared data are under an open source
license and that they adhere to the platform data standards.

### Findability, accessibility, interoperability and reusability

***Why***: Open data has been propelled to the forefront of scientific research
in an era of open, collaborative and reproducible science. By moving towards
large scale, cross-disciplinary research and management projects, there is a
growing need to increase the efficiency of data discovery, access,
interoperability and analysis, for both humans and computationally automated
processes [@reichman2011; @wilkinson2016]. Embracing open-data accelerates research
outcomes, improves transparency and replicability, assists in discovering
erroneous data and corroborating study findings, and decreases overall costs of
research initiatives [@reichman2011; @poisot2013; @lowndes2017]. As such, many
scientific journals (*e.g.* Frontiers, PLoS, Nature, Science), governmental agencies <!-- mettre nom de la revue ou on soumet -->
[*e.g.* @goc2016] and other institutions are adopting open
data-sharing policies and detailed management plans as part of their mandate.
Our goal is to foster efficient and functional open science by creating a
fully open, transparent and replicable open data platform.

@@

By making data and associated tools accessible through a variety of ways, namely
the SLGO web portal (https://ogsl.ca/en), two R packages called *eDrivers*
(https://github.com/orgs/eDrivers/eDrivers) and *eDriversEx*
(https://github.com/orgs/eDrivers/eDriversEx) to access the data through SLGO's
API and to provide analytical tools to explore data, respectively, and a Shiny
application (https://david-beauchesne.shinyapps.io/eDriversApp/) that allows
users to explore drivers data interactively (Figure \ref{diag}).
Note that the data are currently contained within and accessible through the
*eDrivers* R package only, as we are actively working to allow users to
download selected layers from SLGO's web portal and geoserver. The functions
available in *eDrivers* to access the data have however been developed to
ensure forward compatibility once the data migrate to SLGO's geoserver.

@@

By committing to provide an increasingly FAIR infrastructure as we continue
developing it.

@@

***Why***: In the face of uncertainty and in an effort to address impending
environmental changes, adaptive management has been identified as the chief
strategy to guide efficient decision-making
[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
been discussed in the context of multi-drivers and cumulative impact assessments
[@halpern2015; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
truly achieved through a commitment to adaptive monitoring and data reporting
[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
management requires the development of adaptive monitoring tools and
infrastructures, which we seek to address through a continuously-evolving platform.

@@

### Validation

***Why***: Using high-quality data is crucial to ensure robust analyses and
decision-making [@cai2015; @veiga2017]. Data quality assessment (*i.e.* quality
measurement, validation and classification) and management (*i.e.* quality
control and assurance) are therefore crucial steps in data sharing [@veiga2017].

***How***: By selecting a framework to assess and manage data quality and by
providing a data quality report with for all driver layers shared on the
platform.

### Recognition

***Why***: Like peer-reviewed publications, data must also be given its due
importance in scientific endeavors and thus be considered as legitimate citable
products contributing to the overall scientific output of data providers
[@codata2013; @force112014]. Appropriate citations should
therefore be provided for all data layers used and shared by the platform.

***How***: By including the persistent identifier, when possible, and full
citation of all data layers used and shared on the platform.

@@

By adhering to the Data Citation Principles [@force112014], which
focus on citation practices that cover purpose, function and attributes of
citations in an effort to make citations human and machine readable as well as
providing appropriate credit to data products. As such, all data layers shared
on the platform will be identified by a persistent identifier (*e.g.* DOI),
possess a license outlining conditions of use and a clear *how to cite* label
including the authors, the data version and year, an informative title, the
persistent identifier and how to access the data (*e.g.* url).

# St. Lawrence drivers

## Data

The list of drivers for which we sought data was informed by global cumulative
impacts assessments initiatives [@halpern2008; @halpern2015], regional holistic
evaluation of the state of the St. Lawrence [@dufour2007; @benoit2012], and
through communications with regional experts. Drivers are varied in origin,
from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
anthropogenic (*e.g.* fisheries) to natural (*e.g.* temperature anomalies).
Through the data gathering process, we developed and continue developing
collaborations with regional experts and data holders.
We also use global data from seminal works on cumulative impacts on global
habitats [@halpern2008; @halpern2015] and available on the National Center for
Ecological Analysis and Synthesis (NCEAS) online data repository
[Table 1; @halpern2015b]. We selected global data that
were unavailable at the regional scale and that were available at a resolution
adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).

The platform currently provides access to 22 driver layers (Table 1) and we are
actively working on updating or developing additional driver layers.
We divided drivers into 4 groups: coastal, climate, fisheries and marine traffic
related drivers (Table 1). All data layers and methodologies used to produce
driver layers are described in the supplementary materials of this paper and on
the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/).
Also note that the specific
methodologies underlying the figures presented in the paper are described in
the figure captions to ease the flow of text.

## Cumulative exposure

Apart from the northeastern Gulf, cumulative exposure (*i.e.* sum of normalized
driver intensity) is ubiquitous in the St. Lawrence (Figure \ref{footprint}).
Cumulative exposure is generally highest along the coast (Figure \ref{footprint}),
with most intense hotposts (*i.e.* areas of overlapping drivers at high relative
intensity) located in the vicinity of coastal cities (Figure \ref{hotspot}).
In general, offshore areas are less exposed to cumulative drivers, with the
Estuary and the Anticosti Gyre being notable exceptions (Figures \ref{footprint}
and \ref{hotspot}). That is however not to say that offshore areas are free of
exposure, as most of the St. Lawrence is exposed to multiple drivers
(Figures \ref{footprint} and \ref{hotspot}).

>>>>

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
author: David Beauchesne^1,2,\*^, Rémi Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Clément Chion^6^, Peter S. Galbraith^7^, Benjamin Halpern^8,9,10^, Camille Lavoie^2^, Christopher W. McKindsey^7^, Alfonso Mucci^11^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
date: 2018-11-19
fontsize: 12pt
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    fig_width: 3
  pdf_document:
    toc: false
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
relativeurls: true
bibliography: platformMS.bib
csl: frontiers.csl
link-citations: yes
---

@@

\lstset{language=R,
    basicstyle=\small\ttfamily,
    stringstyle=\color{DarkGreen},
    otherkeywords={fetchDrivers, install_github, getData, kde2d},
    morekeywords={TRUE,FALSE},
    deletekeywords={data,frame,length,as,character},
  keywordstyle=\color{blue},
  commentstyle=\color{DarkGreen},
}

<!--
setwd('/users/davidbeauchesne/dropbox/phd/phd_obj1/stressor_spat_dist/platformMS/')
rmarkdown::render('./platformMS.rmd', 'html_document')
rmarkdown::render('./platformMSv2.rmd', 'pdf_document')
rmarkdown::render('./platformMSv2.rmd', 'word_document')
-->

^1^Institut des sciences de la mer, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^2^Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
^4^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^5^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
^6^Institut des sciences de la forêt tempérée, Université du Québec en Outaouais, Gatineau, QC, Canada \newline
^7^Fisheries and Oceans Canada, Maurice Lamontagne Institute, Mont-Joli, QC, Canada \newline
^8^National Center for Ecological Analysis and Synthesis, University of California, Santa Barbara, CA, United States \newline
^9^Bren School of Environmental Science and Management, University of California, Santa Barbara, CA, United States \newline
^10^Imperial College London, Ascot, United Kingdom \newline
^11^Department of Earth & Planetary Sciences, McGill University, Montréal, QC, Canada \newline

**Correspondence**: \newline
David Beauchesne \newline
david.beauchesne@hotmail.com \newline

**Number of figures**: 3 \newline
**Number of tables**: 1 \newline

@@

# Abstract
200 words

**Keywords: open data, St. Lawrence, environmental drivers, holistic, ** up to 8, 5 minimum

# Context

@@

Gathering environmental data for large scale, systematic initiatives can however
be a very challenging and time consuming - not to say painful - process. On one
hand, there is an overwhelming and expanding wealth of data available. Such
information overload may reduce ability to take decision based on scientific
information, massive effort duplication, disproportionate appropriation of
research funds, and it may obscure knowledge gaps amid a sea of information
[@eppler2004]. On the other hand, crucial data are lacking and remain largely
unavailable or inaccessible for a variety of reasons, including proprietary
rights, lack of organizational time, capacity and training, and unwillingness
to share, curtailing our ability for appropriate decision-making. When coupled
with political intricacies, such as multi-jurisdictional management, the data
gathering process can quickly become intractable.

Additionally, the current digital infrastructure is highly
decentralized and the practices are heterogeneous, preventing us from maximizing
benefits from research investments [@wilkinson2016]. Per @wilkinson2016, *"the
reason that we often need several weeks (or months) of specialist technical
effort to gather the data necessary to answer such research questions is not
the lack of appropriate technology; the reason is, that we do not pay our
valuable digital objects the careful attention they deserve when we create and
preserve them"*.

There now exists multiple initiatives that address this issue by assembling,
organizing and sharing vast arrays of environmental data.
Biotic data can be accessed through web portals such as the Ocean Biogeographic
Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
similarly be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
environmental data for specific areas also exist, such as the St. Lawrence
Global Observatory (SLGO; https://ogsl.ca/en), the European Marine Observation
and Data Network (EMODnet; http://www.emodnet.eu/) and the U.S. Integrated Ocean
Observing System (IOOS; https://ioos.noaa.gov/). Essential environmental
parameters are also organized, coordinated and acquired through global
initiatives like the Group on Earth Observations Biodiversity Observation
Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
and the Global Ocean Observing System (GOOS; http://www.goosocean.org).

Environmental drivers are however conspicuously missing from most of these
initiatives [but see @halpern2015], even though high resolution data on their
distribution and intensity is essential to environmental management and impact
assessments. For this paper, we define environmental drivers (hereafter referred
to as drivers) as any process that may disturb a natural system. Drivers may be
natural (e.g. sea surface temperature anomalies and hypoxia) or anthropogenic
(e.g. fisheries and marine pollution) in origin.

To address this issue, we are launching an initiative called *eDrivers*, an
open data platform focused on the description of the spatial distribution and
intensity of drivers in the Estuary and Gulf of St. Lawrence in Canada, a
unique and complex social-ecological system. The overarching goal of this
platform is to create an expert community committed to structuring, sharing
and standardizing drivers data in support to science and management. The
objective of this paper is to present the guiding principles underlying the
inception *eDrivers*, the structure of the platform and potential applications.
Most importantly, we wish to initiate a call for collaboration to any person,
group or organization who might be interested in contributing to this
initiative.

@@

The St. Lawrence, formed by one of the largest estuaries in North America and a
vast interior sea, is an ecologically complex system characterized by highly
variable environmental conditions and oceanographic processes, both in space
and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique and
heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. It concomitantly provides
a wealth of ecosystem services that have historically and contemporarily
benefited the Canadian economy. It holds a rich fisheries industry targeting
over 50 species, serves as the gateway to eastern North-America by granting
access to more than 40 ports, has a booming tourism industry and an expanding
aquaculture production, develops emerging activities and a yet untapped
hydrocarbon potential [@beauchesne2016]. With major investments recently made
and more forthcoming in economic and infrastructure development and research
[*e.g.* @goq2015; @rqm2018], an intensification of the human footprint is
expected in the St. Lawrence.

Arising from this expected intensification and the experience of past and
current ecological tragedies [*e.g.* the collapse of cod fisheries and the
decline of the beluga population; @frank2005; @plourde2014; @dempsey2018] is a
call for an integrative and holistic approach to manage the St. Lawrence as a
whole. Yet, jurisdiction over the St. Lawrence is divided between the Canadian
federal government and five provincial governments, i.e. Newfoundland and
Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec, with
each governmental entity further divided into multiple departments, each with
different mandates. Researchers from dozens of academic institutions and
countless local organizations also share an interest in the St. Lawrence. This
wide range of actors and interests results in highly scattered data management
and sharing processes, which we wish to address through this open data platform
initiative.

@@

Four guiding principles motivate the development of the platform and shaped how
we structured the platform (Figure \ref{diag}).

@@

***Why***: Operating over such large scales in time, space and subject matter
requires a vast and diverse expertise that cannot possibly be possessed by any
one individual or organization. Consequently, we envision a crowd science-like
initiative [see @franzoni2014] that seeks to mobilize all entities
with relevant expertise.

***How***:
By actively inviting, seeking, and developing collaborations as well as
encouraging constructive criticism from the inception and throughout the
lifetime of the platform.

By inviting external community contributions (Figure \ref{diag}).
External researchers or entities wishing to submit data will be invited to do
so mainly through the SLGO web portal. However, submissions through other data
portals will be accepted either through the development of data sharing
agreements or with the caveat that shared data are under an open source license
and that they adhere to the platform data standards.

By promoting and consolidating existing and highly
valuable environmental initiatives already in place in the St. Lawrence. Notable
examples are the annual review of physical [@galbraith2018], chemical, and
biological [@blais2018] oceanographic conditions in the St. Lawrence, the
fisheries monitoring program [@dfo2016], the annual groundfish and shrimp
multidisciplinary survey [@bourdages2017], the characterisation of benthic
[@dutil2011], epipelagic and coastal [@dutil2012] habitats of the St. Lawrence,
and Canada's shoreline classification [@eccc2018]. There are also nascent
efforts to make available information on several human activities in the
St. Lawrence such as the Marine Spatial Data Infrastructure portal, which
provides data on zoning, shipping, port activities, and other human activities
in Canadian waters, including the St. Lawrence [@goc2018].
<!-- include MDDELCC? -->

### Findability, accessibility, interoperability and reusability
<!-- Original: Openness, transparency and replicability -->

***Why***: Open data has been propelled to the forefront of scientific research
in an era of open, collaborative and reproducible science. By moving towards
large scale, cross-disciplinary research and management projects, there is a
growing need to increase the efficiency of data discovery, access,
interoperability and analysis, for both humans and computationally automated
processes [@reichman2011; @wilkinson2016]. Embracing open-data accelerates research
outcomes, improves transparency and replicability, assists in discovering
erroneous data and corroborating study findings, and decreases overall costs of
research initiatives [@reichman2011; @poisot2013; @lowndes2017]. As such, many
scientific journals (*e.g.* PLoS, Nature, Science), governmental agencies
[*e.g.* @goc2016] and other institutions are adopting open
data-sharing policies and detailed management plans as part of their mandate.
Our goal is to foster efficient and functional open science by creating a
fully open, transparent and replicable open data platform.

@@

By hosting on and developing the platform collaboratively with the St. Lawrence
Global Observatory [SLGO; https://ogsl.ca/en]. The mission of SLGO is to promote
and facilitate the accessibility, dissemination and exchange of official and
quality data and information on the St. Lawrence ecosystem through the networking
of organisations and data holders to meet their needs and those of users, to
improve knowledge and to assist decision-making in areas such as public safety,
climate change, transportation, resources and biodiversity conservation.
The SLGO is also one of three regional association spearheading the Canadian
Integreated Ocean Observing System
(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
which will focus on integrating oceanographic data from multiple sources to
make them accessible to end-usersand to enable the national coordination of
ocean observing efforts by integrating isolated or inaccessible data, and by
identifying gaps or duplications in observations and research efforts.

By making data and associated tools accessible through a variety of ways, namely
the SLGO web portal, two R packages called *eDrivers* (https://github.com/orgs/eDrivers/eDrivers)
and *eDriversEx* (https://github.com/orgs/eDrivers/eDriversEx) to access the data trough SLGO's API and to provide
analytical tools to explore data, respectively, and a Shiny application
(https://david-beauchesne.shinyapps.io/platformapp/) that allows users to
explore drivers data interactively (Figure \ref{diag}).
Note that the data are currently contained within and accessible through the
*eDrivers* R package only, as we are actively structuring the project on SLGO's
web portal and geoserver. The functions available in *eDrivers* to access the
data have however been developed to ensure forward compatiblity once the data
migrate to SLGO's geoserver.

@@

By commiting to provide an increasingly FAIR infrastructure (*i.e.* a FAIRport)
as we continue developing it.

@@

***Why***: In the face of uncertainty and in an effort to address impending
environmental changes, adaptive management has been identified as the chief
strategy to guide efficient decision-making
[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
been discussed in the context of multi-drivers and cumulative impact assessments
[@halpern2015; @beauchesne2016; @cote2016]. Adaptive management can only be
truly achieved through a commitment to adaptive monitoring and data reporting
[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
management requires the development of adaptive monitoring tools and
infrastructures, which we seek to address for the St. Lawrence through a
continuously-evolving platform.

@@

### Peer-review and data credit

***Why***: Peer-review is a fundamental tenet of science that ensures robustness
and quality of published research and lends credibility to communicated
scientific outputs [@ware2008; @rennie2016]. In an effort to
uphold the highest standards of scientific research, we wish to provide data
layers that have been reviewed, ideally through peer-reviewed
journals. Like peer-reviewed publications, data must also be given its due
importance in scientific endeavors and thus be considered as legitimate citable
products contributing to the overall scientific output of data providers
[@codata2013; @force112014]. Appropriate citations should
therefore be provided for all data layers used and shared by the platform.

***How***:
By clearly identifying peer-reviewed material and associated citation.

By including the persistent identifier, when possible, and full citation of of
all data layers used and shared on the platform.

@@

By adhering to the Data Citation Principles [@force112014], which
focus on citation practices that cover purpose, function and attributes of
citations in an effort to make citations human and machine readable as well as
providing appropriate credit to data products. As such, all data layers shared
on the platform will be identified by a persistent identifier (*e.g.* DOI),
posseess a license outlining conditions of use and a clear *how to cite* label
including the authors, the data version and year, an informative title, the
persistent identifier and how to access the data (*e.g.* url).

## Content

We first identified a list of 34 drivers (Table 1) known to potentially impact
the St. Lawrence system. This list was informed by global cumulative impacts
assessments initiatives [@halpern2008; @halpern2015], regional holistic
evaluation of the state of the St. Lawrence [@dufour2007; @benoit2012], and
through communications with regional experts. Drivers are varied in origin,
from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
anthropogenic (*e.g.* fisheries) to natural (*e.g.* temperature anomalies).

We have developed and continue developing collaborations with regional experts
to gather data and characterize the distribution and intensity of these drivers
in the St. Lawrence system. We have thus far characterized 12 drivers and are
actively working on the characterization of 12 more, with their development at
various stages of completion (Table 1). We are also actively seeking
collaborations or data for 10 drivers (Table 1). We also use global data from
the National Center for Ecological Analysis and Synthesis (NCEAS) online data
repository [Table 1; @halpern2015b], as certain layers provide valuable
information at a resolution adequate for use at regional scales or currently
constitute the best available knowledge for the St. Lawrence (*e.g.* marine
pollution).
<!-- Review with the final list of data that I will indeed be able to share -->

# Applications

Assembling all drivers in a single platform will provide a wide array of
visualization and analytical possibilities that can be useful for both decision
making and scientific research. Here we describe two such potential
applications, *i.e.* multi-drivers assessment and multi-scale management.
Furthermore, we demonstrate the ease with which those applications will be
possible through the platform by providing the code to replicate the elements
presented on \ref{fig1} for multi-drivers assessment in box 1.

## Multi-drivers assessment

This platform initiative will enable usual, single driver impact assessments,
but it will also pave the way towards holistic assessments of the state of the
St. Lawrence by enabling the study of combinations of multiple drivers. For
instance, drivers like hypoxia (Figure \ref{fig1}A) and demersal destructive
fisheries (Figure \ref{fig1}B) are known to occur mainly in deeper areas of the
St. Lawrence, and hence a synergy between the two drivers could be anticipated.
By combining them (Figure \ref{fig1}C), we can evaluate which
regions in the St. Lawrence are more likely to be affected by both drivers
simultaneously and evaluate the relative intensity at which they tend to
co-occur (Figure \ref{fig1}D). Doing this, we observe that hypoxia and bottom
fisheries tend to co-occur in the St. Lawrence and that they do so at relatively
high intensities, with extensively used fishing grounds also characterized by
low levels of oxygen neighboring the limits of hypoxia (*i.e.* ~0.7 mg*L^-1^,
corresponding to ~30% saturation).
